---
title: "Unit 11 For Live Session"
author: "Chad Madding"
date: "March 15, 2020"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = "show")
library(readr)
library(tswge)
library(tseries)
library(forecast)
```

## Unit 11 For Live Session

**In preparation for the live session, please complete the following. Be sure and submit your work to the "Unit 11: "For Live Session" Assignment" assignment on 2DS:**  

Please address each activity on at least one PowerPoint slide and submit via the online campus.
Our purpose is to update the Sunspot analysis. Download the most current yearly mean sunspot data and with this data:  

####1. Plot the data.  
```{r, echo=FALSE, results='hide'}
SS = read_delim("Sunspot Numbers Thru 2019.csv", 
    ";", escape_double = FALSE, trim_ws = TRUE)

#Convert to a time series
SS = ts(SS$SunSpots, start = 1700, end = 2019)
#A simple plot of the data
plot(SS)
#Plot Data, Sample Autocorrelations, Periodogram, and Parzen Spectral Estimate
plotts.sample.wge(SS, arlimits = TRUE)
```

####2. Comment on its stationarity.  
Sample autocorrelations damp quickly and seem to indicate stationarity. A unit root test also shows stationarity. With a frequency peek at 0.0937 there does seem to be some seasonality in the data (1/0.937=10.67).  
```{r stationarity, echo=FALSE, results='hide'}
acf(SS)
pacf(SS)
```
```{r}
adf.test(SS)
```

####3. Use aic5.wge to estimate the p and q of the model. You may use your choice of AIC/AICC/BIC.  
```{r aic5}
#Checking the AIC options
aic5.wge(SS,p=0:15,q=0:2,type="aic")
aic5.wge(SS,p=0:15,q=0:2,type="aicc")
aic5.wge(SS,p=0:15,q=0:2,type="bic")
```
All the AIC's point to and AR(9).  

```{r Forcast package}
#playing around with the Forcast package to see what it picks
#differencing needed to stationarize the series
ndiffs(SS)
SSAA=auto.arima(SS,stepwise = FALSE,  approximation=FALSE, seasonal = TRUE, stationary = TRUE)
SSAA
forecast(SSAA, h=15)
```

####4. Fit the model using your model identification (p and q). You may use any of the estimates you like (maximum likelihood, Yule-Walker, Burg).  
I am choosing to go with the AR(3) model from the auto.arima.  
```{r Fit the model}
#Maximum Likelihood Estimation
SS3.mle=est.ar.wge(SS,p=3,type='mle')
mean(SS) #[1] 78.735
SS3.mle$avar #[1] 644.4032
#Yule Walker Estimate
#SS.yw=est.ar.wge(SS,p=9,type='yw')
#Burg Estimates
#SS.burg=est.ar.wge(SS,p=9,type='burg')
```

** Final AR(3) model**  

$$(1-1.28B+0.49B^2+0.14B^3)(x_t-78.74)=a_y\  \sigma^2_a=644.4$$

####5. Use this model to generate an ASE from forecasting the last 15 years of sunspot data.  

The maximum likelihood scored just a bit better so I decided to stay with it's forecast.  

```{r generate an ASE}
SS3.mle.est = fore.arma.wge(SS, phi = SS3.mle$phi, theta = 0,lastn = T, n.ahead = 15)
SS3.mle.ase = mean((SS[(320-15+1):320] - SS3.mle.est$f)^2)
SS3.mle.ase
#SS.mle.est = fore.arma.wge(SS, phi = SS.mle$phi, theta = 0,lastn = T, n.ahead = 15)
#SS.mle.ase = mean((SS[(320-15+1):320] - SS.mle.est$f)^2)
#SS.mle.ase
#SS.yw.est = fore.arma.wge(SS, phi = SS.yw$phi, theta = 0,lastn = T, n.ahead = 15)
#SS.yw.ase = mean((SS[(176-15+1):176] - SS.yw.est$f)^2)
#SS.yw.ase
#SS.burg.est = fore.arma.wge(SS, phi = SS.burg$phi, theta = 0,lastn = T, n.ahead = 15)
#SS.burg.ase = mean((SS[(176-15+1):176] - SS.burg.est$f)^2)
#SS.burg.ase
```

####6. Now fit a seasonal model to the Sunspot data (you pick the value of s), and find the ASE for this model using the last 15 years of sunspot data.  

**Non Stationary**  

```{r}
SS.overfit=est.ar.wge(SS,p = 12) #overfit for seasonality
```

```{r known 11 table}
#(1-B^11)
factor.wge(c(rep(0,11),1))
```
#### Transform  

Base on the overfit table we can now transform the data to remove the seasonality.  

Transform data to create $Y_t=(1-B^{11})X_t$  

```{r Transform data s 11}
#rep will repeat the 11 zeros then we can add a one at the end
SS_11 = artrans.wge(SS,phi.tr = c(rep(0,10),1))
```

The transformed data appear stationary, so we use AIC to identify a model.

We use **aic5.wge** to model the transformed data  
When using AIC to model data that has been stationarized using the seasonal transform $(1-B^{11})$, it is good practice to allow a range of p values to include s to uncover any seasonal stationary information that might be in the data.

```{r}
aic5.wge(SS_11)
aic5.wge(SS_11, type = "bic")
```

**Seasonal ARMA(2,0) s = 11**  

AIC selects an AR(2,0) model. Factoring the ARMA(2,0) model we obtain:  

```{r est the y}
AR2 = est.arma.wge(SS_11,p = 2,q = 0)
```

```{r transform and check whitness}
SS_11_AR2 = artrans.wge(SS_11,AR2$phi)
ljung.wge(SS_11_AR2)
ljung.wge(SS_11_AR2, K = 48)
```

```{r forcast with s11}
f = fore.aruma.wge(SS,s = 11,phi = AR2$phi,n.ahead = 15,limits = F,lastn = TRUE)

SS_11_AR2.ASE = mean((SS[(length(SS)-14):length(SS)] - f$f)^2)

SS_11_AR2.ASE
```

####7. Describe which model you prefer, and why.  

The best model turned out to be the AR(3) that auto.arima from the Forecast package picked. The ASE was also lower than the seasonal model. This was a stationary model to start with and that could have lead to the better performance.  

####8. Use your best model (the one you choose) to forecast the next 10 years of sunspot data.  

**Final AR(3) model**  
  
$$(1-1.28B+0.49B^2+0.14B^3)(x_t-78.74)=a_y\  \sigma^2_a=644.4$$  

```{r Final Forecasts}
SS3.f = fore.arma.wge(SS,phi = SS3.mle$phi,theta = 0, n.ahead = 10,limits = F, lastn = F)
```

#### Analysis of AcuSpike Web Hits

Given what you have learned so far, model the Accuspike web page hits data, and provide the desired forecasts.  
This does not need to be a formal analysis with a write-up.  
You should be able to explore the data, identify a model, estimate the model, judge the appropriateness of the model, and provide the forecasts.  

#### Plot the Data
```{r read in accuspike}
AS = read.csv("accuspike.csv",header = TRUE)
#Convert to a time series
AS = ts(AS$Active.Users)
```

```{r}
#A simple plot of the data
plot(AS)
#Plot Data, Sample Autocorrelations, Periodogram, and Parzen Spectral Estimate
plotts.sample.wge(AS, arlimits = TRUE)
```

Sample autocorrelations damp quickly and seem to indicate stationarity. A unit root test also shows stationarity. With a frequency peek around 0.15 (1/0.15=6.67) there could be some weekly seasonality in the data.  

```{r AS stationarity, echo=FALSE, results='hide'}
acf(AS)
pacf(AS)
```

```{r}
adf.test(AS)
```

```{r AS aic5}
#Checking the AIC options
aic5.wge(AS)
```

```{r accuspike Forcast package}
#Use the Forcast package to see what it picks
#differencing needed to stationarize the series
ndiffs(AS)
ASAA=auto.arima(AS,stepwise = FALSE,  approximation=FALSE)
ASAA
ASAA.f=forecast(ASAA, h=14)
```

The forecast package picked a ARMA(4,1) and that matches one to the AIC5's so we will be using the (4,1).
```{r AS Fit the model}
#Fit with 4,1
AS41=est.arma.wge(AS,p=4,q=1,factor = TRUE)
AS41$phi #[1]  1.3614265 -0.5113544 -0.1158767  0.2080940
AS41$theta #[1] 0.6585371
mean(AS) #[1] 80.35
AS41$avar #[1] 386.29
```

**Final ARMA(4,1) model**  

$$(1-1.36B+0.51B^2+0.12B^3-0.21B^4)(x_t-80.35)=(1-0.66B)a_y\  \sigma^2_a=386.29$$
Forecast out two weeks

```{r AS Final Forecasts}
AS41.f = fore.arma.wge(AS,phi = AS41$phi,theta = AS41$theta, n.ahead = 14,limits = F, lastn = F)
```

```{r compare the 2 forcast}
AS41.f$f
ASAA.f
plot(ASAA.f)
```
