---
title: "Using Turnstile Data To Forecast Student Worker Staffing"
author: "Chad Madding, Shane Weinstock"
date: "April 11, 2020"
output:
  html_document: default
  word_document: default
---

### Clients  

This report is being prepared for the management of the Dedman Center for Lifetime Sports at Southern Methodist University.  

### Directive  

We have been tasked with providing general usage of the facilities in the form of entry data. The main goal will be to use ID swipes as a response variable to assist in hiring patters for the facilities’ student staff. It is the goal of the report to assist in the spotting of trends and thus correctly identify staffing needs. Forecasting and general knowledge about building usage will also assist with budgeting needs. 

### Data  

The data provided to us has been collected from three turnstiles at the Dedman Center for Lifetime Sports on the campus of Southern Methodist University. The data set is a record of the time of swipe, the turnstile used, and an anonymized student ID number. The data was collected from January 2nd, 2019 through March 11th, 2020, and consists of 414,156 entries. Entry and error swipes are included in the data. It should be noted that the turnstiles are only used to enter the facility, and no swipe is required to exit the building. We have also collected hourly weather data from NOAA for the same period to assist in forecasting.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(readxl)
library(tswge)
library(plyr)
library(dplyr)
library(forecast)
library(ggplot2)
library(ggthemes)
library(nnfor)
```

```{r Read in, echo=FALSE, include=FALSE}
#Read in the data
SMUSwipe <- read_excel("AnonDataTurnstile.xlsx")
weather <- read_csv("2085029.csv")
#pull out just the hourly weather data
HourlyWeather = weather[ , c(2, 42:57)]
```

```{r Cleaning, echo=FALSE}
#Data Cleaning
#Remove any rejected \ change state rows
SMUSwipe = SMUSwipe[SMUSwipe[,2]=="CardAdmitted",]
#Create a Time column
SMUSwipe$Time <- format(SMUSwipe$LDT,"%H:%M:%S")
#Create a Day column
SMUSwipe$Day <- format(SMUSwipe$LDT,"%d")

#Create a Date column
SMUSwipe$Date <- as.Date(SMUSwipe$LDT)
#Create a Day coloum
SMUSwipe$Day <- weekdays(as.Date(SMUSwipe$LDT))
#Create an Hours coloum for count
SMUSwipe$Hours <- format(SMUSwipe$LDT,"%Y-%m-%d %H")
#Create an Month coloum for count
SMUSwipe$Month <- format(SMUSwipe$LDT,"%Y-%m")
#Create an Hour coloum for count
SMUSwipe$Hour <- format(SMUSwipe$LDT,"%H")
```

To get ready for merging the temperature data we needed to clean up the hours in the SMU data.
```{r Hour breakdown SMUSwipe}
#Group hours
SMUSwipe$TempTime <- format(SMUSwipe$LDT,"%Y-%m-%d")
SMUSwipe$Hour <-ifelse(SMUSwipe$Hour <= '04','04',
                      ifelse(SMUSwipe$Hour <= '08','08',
                      ifelse(SMUSwipe$Hour <=  12, 12,
                      ifelse(SMUSwipe$Hour <=  16, 16,
                      ifelse(SMUSwipe$Hour <=  20, 20,
                      ifelse(SMUSwipe$Hour <=  24, 24))))))
SMUSwipe$TempTime<-paste(SMUSwipe$TempTime, SMUSwipe$Hour, sep=" ")
```

We made the data the same format for the weather data as well as in the SMU information.
```{r Hour breakdown HourlyWeather}
#Create a TempTime in HourlyWeather to merge Temperature data
HourlyWeather$TempTime <- format(HourlyWeather$DATE,"%Y-%m-%d")
HourlyWeather$Hour <- format(HourlyWeather$DATE,"%H")

HourlyWeather$Hour <-ifelse(HourlyWeather$Hour <= '04','04',
                      ifelse(HourlyWeather$Hour <= '08','08',
                      ifelse(HourlyWeather$Hour <=  12, 12,
                      ifelse(HourlyWeather$Hour <=  16, 16,
                      ifelse(HourlyWeather$Hour <=  20, 20,
                      ifelse(HourlyWeather$Hour <=  24, 24))))))
HourlyWeather$TempTime<-paste(HourlyWeather$TempTime, HourlyWeather$Hour, sep=" ")
#Recreate the Hour data
SMUSwipe$Hour <- format(SMUSwipe$LDT,"%H")
```
Here we are merging on TempTime to pull in hourly weather data
```{r}
#Merge on TempTime to pull in hourly weather data
SMUSwipe = merge(SMUSwipe, HourlyWeather, by.x='TempTime', by.y='TempTime',all.x = TRUE, all.y = TRUE)
#Remove dup after the left join
SMUSwipe = distinct(SMUSwipe, LDT, .keep_all = TRUE)
SMUSwipe = SMUSwipe[which(!is.na(SMUSwipe$`LDT`)),]

#sum(is.na(SMUSwipe$HourlyDryBulbTemperature))

#Fill in NA's from Temperature data with number close to the mean
SMUSwipe$HourlyDryBulbTemperature[is.na(SMUSwipe$HourlyDryBulbTemperature)] <- 65
SMUSwipe$HourlyAltimeterSetting[is.na(SMUSwipe$HourlyAltimeterSetting)] <- 30
SMUSwipe$HourlyDewPointTemperature[is.na(SMUSwipe$HourlyDewPointTemperature)] <- 49
SMUSwipe$HourlyRelativeHumidity[is.na(SMUSwipe$HourlyRelativeHumidity)] <- 58
SMUSwipe$HourlyWindSpeed[is.na(SMUSwipe$HourlyWindSpeed)] <- 0

#Renaming some information for ease of use
SMUSwipe$Temperature <- SMUSwipe$HourlyDryBulbTemperature
SMUSwipe$Hour <- SMUSwipe$Hour.x

#Dropping some data we no longer need
drop <- c("Message Type","Hour.x","Hour.y", "TempTime", "DATE", "HourlyPressureChange", "HourlyPressureTendency", "HourlySeaLevelPressure", "HourlyWetBulbTemperature", "HourlyDryBulbTemperature")
SMUSwipe = SMUSwipe[,!(names(SMUSwipe) %in% drop)]
```

Create and export datasets for time series studies
```{r}
#Read in the data
SMU = read.csv('SMUSwipe.csv',header = TRUE)
#Look at the top of the data
head(SMU)

#Dropping some data we no longer need
drop <- c("ID.", "interval_15", "Minutes_15","Time")
SMU = SMU[,!(names(SMU) %in% drop)]

SMU$Week<-paste(SMU$Month, SMU$Day, sep=" ")

#Renaming some information for ease of use
#SMUSwipe$Temperature <- SMUSwipe$HourlyDryBulbTemperature
#SMUSwipe$Hour <- SMUSwipe$Hour.x

colnames(SMU)[2] = "Turnstile"

SMU$Date <- as.Date(SMU$LDT)

# Extract day of the week (Saturday = 6)
SMU$Week_Day <- as.numeric(format(format.Date(SMU$LDT,"%w")))

# Adjust end-of-week date (first saturday from the original Date)
SMU$End_of_Week <- SMU$Date + (6 - SMU$Week_Day)

HourSwipes = dplyr::count(SMU,Hours)
DaySwipes = dplyr::count(SMU,Date)
WeekSwipes = dplyr::count(SMU,End_of_Week)
MonthSwipes = dplyr::count(SMU,Month)

HourSwipesTemp = merge(HourSwipes,SMU, by='Hours')
HourSwipesTemp = distinct(HourSwipesTemp, Hours, .keep_all = TRUE)

DaySwipesTemp = merge(DaySwipes,SMU, by='Date')
DaySwipesTemp = distinct(DaySwipesTemp, Date, .keep_all = TRUE)
# Aggregate over Date and Temperature
DaySwipesTemp$DayTemperature = aggregate(Temperature~Date, FUN=mean, data=SMU, na.rm=TRUE)
DaySwipesTemp = DaySwipesTemp[,!(names(DaySwipesTemp) %in% 'Temperature')]

WeekSwipesTemp = merge(WeekSwipes,SMU, by='End_of_Week')
WeekSwipesTemp = distinct(DaySwipesTemp, End_of_Week, .keep_all = TRUE)
# Aggregate over week and Temperature
WeekSwipesTemp$WeekTemperature = aggregate(Temperature~End_of_Week, FUN=mean, data=SMU, na.rm=TRUE)
WeekSwipesTemp = WeekSwipesTemp[,!(names(WeekSwipesTemp) %in% 'Temperature')]

MonthSwipesTemp = merge(MonthSwipes,SMU, by='Month')
MonthSwipesTemp = distinct(MonthSwipesTemp, Month, .keep_all = TRUE)
# Aggregate over week and Temperature
MonthSwipesTemp$MonthTemperature = aggregate(Temperature~Month, FUN=mean, data=SMU, na.rm=TRUE)
MonthSwipesTemp = MonthSwipesTemp[,!(names(MonthSwipesTemp) %in% 'Temperature')]

#Change n to IDSwipes
colnames(HourSwipesTemp)[2] = "IDSwipes"
colnames(DaySwipesTemp)[2] = "IDSwipes"
colnames(WeekSwipesTemp)[2] = "IDSwipes"
colnames(MonthSwipesTemp)[2] = "IDSwipes"

#Dropping some data we no longer need
drop <- c("LDT", "Minutes", "Week", "Turnstile")
HourSwipesTemp = HourSwipesTemp[,!(names(HourSwipesTemp) %in% drop)]
DaySwipesTemp = DaySwipesTemp[,!(names(DaySwipesTemp) %in% drop)]
WeekSwipesTemp = WeekSwipesTemp[,!(names(WeekSwipesTemp) %in% drop)]
MonthSwipesTemp = MonthSwipesTemp[,!(names(MonthSwipesTemp) %in% drop)]

#Export the dataset
write.csv(HourSwipesTemp,"DedmanHourlySwipe.csv", row.names = FALSE)
write.csv(DaySwipesTemp,"DedmanDailySwipe.csv", row.names = FALSE)
write.csv(WeekSwipesTemp,"DedmanWeeklySwipe.csv", row.names = FALSE)
write.csv(MonthSwipesTemp,"DedmanMonthlySwipe.csv", row.names = FALSE)
```



We now have a data set with weather data along with card swipes.
```{r}
head(SMUSwipe)
```

Hourly ID Swipes for all Turnstiles?
```{r Hour Group}
#Group all badge swipes by the hour

hourplot = dplyr::count(SMUSwipe,Hour)

hourplot %>%
  ggplot(aes(x=Hour,y=n, group=1))+
  geom_line()+
  geom_point()+
  theme_economist()+
  scale_colour_economist()+
  ggtitle('Hourly ID Swipes for all Turnstiles')
```

What are the Daily ID Swipes for all Turnstiles?  
```{r}
dayCount=dplyr::count(SMUSwipe, Day)
dayCount$Day <- factor(dayCount$Day, levels= c("Sunday", "Monday", 
    "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

dayCount=dayCount[order(dayCount$Day), ]
  
dayCount %>%
  ggplot(aes(x=Day,y=n, group=1))+
  geom_line()+
  geom_point()+
  theme_economist()+
  scale_colour_economist()+
  theme_economist()+
  ggtitle('Daily ID Swipes for all Turnstiles')+
  ylab('ID Swipes')
```



Look at turnstile information.  
Create individual turnstile data
```{r}
#Create a dataset for each turnstyle
SMUSwipeTurn1 = SMUSwipe[SMUSwipe[,3]=="DEDM (101.1LB)",]
SMUSwipeTurn2 = SMUSwipe[SMUSwipe[,3]=="DEDM (101.2LB)",]
SMUSwipeTurn3 = SMUSwipe[SMUSwipe[,3]=="DEDM (101.3LB)",]
```

```{r}

df = dplyr::count(SMUSwipeTurn3, Hours)

HTurn1 = dplyr::count(SMUSwipeTurn1, Hour)
HTurn1 %>%
  ggplot(aes(x=Hour,y=n, group=1))+
  geom_line()+
  geom_point()+
  theme_economist()+
  scale_colour_economist()+
  ggtitle('Hourly Card Swipes for Turnstile 1')

HTurn2 = dplyr::count(SMUSwipeTurn2, Hour)
HTurn2 %>%
  ggplot(aes(x=Hour,y=n, group=1))+
  geom_line()+
  geom_point()+
  theme_economist()+
  scale_colour_economist()+
  ggtitle('Hourly Card Swipes for Turnstile 2')

HTurn3 = dplyr::count(SMUSwipeTurn3, Hour)
HTurn3 %>%
  ggplot(aes(x=Hour,y=n, group=1))+
  geom_line()+
  geom_point()+
  theme_economist()+
  scale_colour_economist()+
  ggtitle('Hourly Card Swipes for Turnstile 3')
```

Looks like turnstile 3 gets used more than the others but the pattern of usage looks the same across each.  

This data is anamonized but we can still see who are some of the top users.
```{r ID Count}
#This will count up the times users swiped in. 
IDCount = dplyr::count(SMUSwipe, `ID#`)
#Print off the top 10 users ordering nuency in decending order
head(IDCount[order(IDCount$n, decreasing = TRUE),],10)
```
Look at some user data.
```{r}
#Breakdown user numbers
summary(IDCount$n)
#how many users
dplyr::count(IDCount)
```
