---
title: 'Unit 13: For Live Session'
author: "Chad Madding"
date: "March 29, 2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GGally)
library(nnfor)
library(readr)
library(tswge)
library(vars)
library(forecast)
```

## Unit 13: For Live Session

In preparation for the live session, please complete the following and submit your PowerPoint. Be sure and submit your work to the "Unit 13: "For Live Session" Assignment" assignment on 2DS.

#### 1. Model the sunspot/melanoma data with a MLP. Use the model to provide eight-step-ahead forecasts. Finally, calculate the ASE, and compare your NN model to the VAR model we estimated previously.  

```{r #1 SM}
#Read in the data
SM = read.csv('Melanoma_Sunspot.csv',header = TRUE)
SMsmall = SM[1:29,]

SMsmallDF = data.frame(Sunspot = ts(SMsmall$Sunspot))
fit.mlp = mlp(ts(SMsmall$Melanoma),reps = 50,comb = "mean",xreg = SMsmallDF)
fit.mlp
plot(fit.mlp)
SMDF = data.frame(Sunspot = ts(SM$Sunspot))
fore.mlp = forecast(fit.mlp, h = 8, xreg = SMDF)
plot(fore.mlp)

plot(SM$Melanoma, type = "l")
lines(seq(30,37,1),fore.mlp$mean,col = "blue")

ASE_SM_NN = mean((SM$Melanoma[30:37] - fore.mlp$mean)^2)
ASE_SM_NN
```

##### Sunspot/Melanoma VAR Model estimated previously

**VAR Model Forecasts Seasonally Differenced Data**  

```{r}
SM = read.csv('Melanoma_Sunspot.csv',header = TRUE)
SMsmall = SM[1:29,]

VAR_SM = VAR(cbind(SMsmall$Melanoma,SMsmall$Sunspot),lag.max = 5, type = "both")

pred = predict(VAR_SM,n.ahead = 8)

plot(SM$Melanoma, type = "l")
lines(seq(30,37,1),pred$fcst$y1[,1],col = "red")

ASE_SM_VAR = mean((SM$Melanoma[30:37] - pred$fcst$y1[1:8])^2)
ASE_SM_VAR
```

#### 2. Model the Schumway pollution data with a MLP model, and provide 10-step-ahead forecasts. Calculate the ASE, and add your analysis to a slide in your Unit 13 For Live Session PowerPoint deck.  

MLP MODEL FOR CARDIAC MORTALITY DATA
```{r #2 cmort}
#Forcast out?
CMForcast = 10

#Read in the data
CM = read.csv('LA_Cmort_Study.csv',header = TRUE)

#Lag 1
CM$temp_1 = dplyr::lag(CM$temp,1)

#Look at the top of the data
head(CM)

ccf(CM$cmort, CM$temp)

#get the total rows
CMRows = nrow(CM)

CMsmall = CM[2:(CMRows-CMForcast),]

CMsmallDF = data.frame(Week = ts(CMsmall$Week),temp = ts(CMsmall$temp), part = ts(CMsmall$part),temp_1 = ts(CMsmall$temp_1))

fit.mlp1 = mlp(ts(CMsmall$cmort),reps = 50,comb = "mean",xreg = CMsmallDF)
fit.mlp1
plot(fit.mlp1)
CMDF = data.frame(Week = ts(CM$Week),temp = ts(CM$temp), part = ts(CM$part), temp_1 = ts(CM$temp_1))

fore.mlp1 = forecast(fit.mlp1, h = CMForcast, xreg = CMDF)

plot(fore.mlp1)

ASE_CM_NN = mean((CM$cmort[((CMRows-CMForcast)+1):CMRows] - fore.mlp1$mean)^2)
ASE_CM_NN

#while (!is.null(dev.list()))  dev.off()
#par(mfrow = c(2,1))

#Plot
plot(seq(1,CMRows,1), CM$cmort, type = "l",xlim = c(0,CMRows), ylab = "Cardiac Mortality", main = paste(CMForcast," Week Cardiac Mortality Forecast"))
lines(seq((CMRows-CMForcast)+1,CMRows,1), fore.mlp1$mean, type = "l", col = "blue")
```

VAR Model 3 seasonal with Lag 1 Temp
```{r #2 cmort VAR}

#Lag the temp data one
CM$temp_1 = dplyr::lag(CM$temp,1)
ggpairs(CM)

VARselect(cbind(CM$cmort[2:508], CM$part[2:508], CM$temp[2:508]),lag.max = 10, season = 52, type = "both")

#VAR with p = 2 (Integer for the lag order (default is p=1))
CMortVAR = VAR(cbind(CM$cmort[2:508], CM$part[2:508], CM$temp[2:508]),season = 52, type = "both",p = 2)
preds=predict(CMortVAR,n.ahead=30)

#Plot
plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,528), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(509,538,1), preds$fcst$y1[,1], type = "l", col = "red")


#Find ASE using last 30
CMsmall = CM[1:478,]

#Start and 2 since the lagged variable is NA at first index
VARselect(cbind(CMsmall$cmort[1:478], CMsmall$part[1:478], CMsmall$temp[1:478]),lag.max = 10, season = 52, type = "both")

CMortVAR = VAR(cbind(CMsmall$cmort[1:478], CMsmall$part[1:478], CMsmall$temp[1:478]),season = 52, type = "both",p = 2)
preds=predict(CMortVAR,n.ahead=30)


#while (!is.null(dev.list()))  dev.off()
#par(mfrow = c(2,1))

#Plot
plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,508), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(479,508,1), preds$fcst$y1[,1], type = "l", col = "red")

#Plot
plot(seq(479,508,1), CM$cmort[479:508], type = "l",xlim = c(479,508), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(479,508,1), preds$fcst$y1[,1], type = "l", col = "red")


ASE_CM_VAR = mean((CM$cmort[479:508] - preds$fcst$y1[,1])^2)
ASE_CM_VAR
```

#### 3. Model your time series with a MLP, and compare it with ARMA/ARIMA models you have fit to it in the past.

```{r #3}
#Forcast out?
SMUForcast = 48

#Read in the data
SMU = read.csv('DedmanHourleySwipe.csv',header = TRUE)
#Look at the top of the data
head(SMU)

plotts.sample.wge(SMU$IDSwipes)

#Just using Feb. 11th to March 11th 2020 data
SMU = SMU[5891:6360,]

#get the total rows
CMRows = nrow(SMU)

SMUsmall = SMU[1:(CMRows-SMUForcast),]

ggpairs(SMU[c(2,10,17,18)])

SMUsmallDF = data.frame(Temperature = ts(SMUsmall$Temperature), Humidity = ts(SMUsmall$HourlyRelativeHumidity),Hour = ts(SMUsmall$Hour))

fit.mlp = mlp(ts(SMUsmall$IDSwipes),reps = 20,comb = "mean",xreg = SMUsmallDF)

fit.mlp
plot(fit.mlp)

SMUDF = data.frame(Temperature = ts(SMU$Temperature), Humidity = ts(SMU$HourlyRelativeHumidity), Hour = ts(SMU$Hour))

fore.mlp = forecast(fit.mlp, h = 48 , xreg = SMUDF)

fore.mlp

plot(fore.mlp)

ASE_SMU_NN = mean((SMU$IDSwipes[((CMRows-SMUForcast)+1):CMRows] - fore.mlp$mean)^2)
ASE_SMU_NN

while (!is.null(dev.list()))  dev.off()
par(mfrow = c(2,1))

#Plot
plot(seq(1,CMRows,1), SMU$IDSwipes, type = "l",xlim = c(0,CMRows), ylab = "ID Swipes", main = paste(SMUForcast," Hour Turnstile Forecast"))
lines(seq(((CMRows-SMUForcast)+1),CMRows,1), fore.mlp$mean, type = "l", col = "red")

#Plot
plot(seq(((CMRows-SMUForcast)+1),CMRows,1), SMU$IDSwipes[((CMRows-SMUForcast)+1):CMRows], type = "l",xlim = c((CMRows-SMUForcast)+1,CMRows), ylab = "ID Swipes", main = paste(SMUForcast," Hour Turnstile Forecast"))
lines(seq(((CMRows-SMUForcast)+1),CMRows,1), fore.mlp$mean, type = "l", col = "red")
```

SMU data NN using the deafult settings.
```{r SMU Dedman default}
fit.mlp = mlp(ts(SMUsmall$IDSwipes))
fit.mlp
plot(fit.mlp)
fore.mlp = forecast(fit.mlp, h = 48)
plot(fore.mlp)

ASE_SMU_NNid = mean((SMU$IDSwipes[((CMRows-SMUForcast)+1):CMRows] - fore.mlp$mean)^2)
ASE_SMU_NNid

```

VAR for Dedman data
```{r}
#VAR
SMForcast = 48
SM = read.csv('DedmanHourleySwipe.csv',header = TRUE)
#Just using Feb. 11th to March 11th 2020 data
SM = SM[5891:6360,]
#Just using March 1st to March 11th 2020 data
#SM = SM[6189:6360,]
#get the total rows
SMRows = nrow(SM)
SMsmall = SM[1:(SMRows-SMForcast),]

#VAR Model 3 seasonal with Lag 1 Temp
SMsmall$temp_1 = dplyr::lag(SMsmall$Temperature,1)

VARselect(cbind(SMsmall$IDSwipes[1:(SMRows-SMForcast)], SMsmall$Temperature[1:(SMRows-SMForcast)], SMsmall$HourlyRelativeHumidity[1:(SMRows-SMForcast)],SMsmall$Hours[1:(SMRows-SMForcast)]))

VAR_SM = VAR(cbind(SMsmall$IDSwipes[1:(SMRows-SMForcast)], SMsmall$Temperature[1:(SMRows-SMForcast)], SMsmall$HourlyRelativeHumidity[1:(SMRows-SMForcast)],SMsmall$Hours[1:(SMRows-SMForcast)]))

pred = predict(VAR_SM,n.ahead = SMForcast)

#while (!is.null(dev.list()))  dev.off()

plot(SM$IDSwipes, type = "l")
lines(seq((SMRows-SMForcast)+1,SMRows,1),pred$fcst$y1[,1],col = "red")

ASE_SMU_VAR = mean((SM$IDSwipes[((SMRows-SMForcast)+1):SMRows] - pred$fcst$y1[1:SMForcast])^2)

ASE_SMU_VAR
```

Nottingham data for use with NN
```{r}
#Average Monthly Temperatures at Nottingham, 1920–1939
data("nottem")
nottem = ts(nottem, start = c(1920,1), end = c(1939,12),frequency = 12)
nottem
```
Nottingham data NN using the deafult settings.
```{r NOT NN default}
fit.mlp.not = mlp(nottem)
fit.mlp.not
plot(fit.mlp.not)
fore.mlp.not = forecast(fit.mlp.not, h = 48)
plot(fore.mlp.not)

ASE_NOT_NNid = mean((nottem[((240-48)+1):240] - fore.mlp.not$mean)^2)
ASE_NOT_NNid

```
