---
title: 'Unit 10: For Live Session'
author: "Chad Madding"
date: "March 9, 2020"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tswge)
library(readr)
```
### Select a data set, and complete one of each model ID below:
#### ARIMA

First we read in the data collected from:
https://archive.ics.uci.edu/ml/datasets/Appliances+energy+prediction
```{r read}
# read in the data
x = read.csv("energydata_complete.csv", header = TRUE)
```

Convert the data to time series
```{r convert to TS}
#Just looking at the Appliances data
x=ts(x$Appliances)
```

Plot the data and see the associated (damping) sample autocorrelations
```{r Plot, echo=FALSE}
#Plot the data
plot.x=plotts.sample.wge(x)
plot.x
```

We will follow the Classical Box-Jenkins Procedure for Including a Unit Root in the Model.
If data are wandering and sample autocorrelations damp slowly difference the data with a d-1
```{r difference}
# difference the data
x.d1=artrans.wge(x,phi.tr=1)
```

The data looks stationary so there is no need to difference again.
Estimate the parameters of the "stationarized data".
```{r AIC5 estimate}
# x.d1 appears to be stationary
aic5.wge(x.d1,p=0:5,q=0:2)
```

AIC picks an ARMA(3,1)
```{r estimate}
# AIC picks an ARMA(3,1)
# which seems reasonable from
est.x.d1=est.arma.wge(x.d1,p=3,q=1)
```

Pull out what will make up the final model
```{r final model}
#Pull out what will make up the final model
est.x.d1$phi
est.x.d1$theta
est.x.d1$avar
mean(x)
#(1_B)^2(1-1.27B+.68B^2)(Xt-1512)=at WHV=1.03
```
** Final Model**
$$(1-B)(1-0.731B+0.203B^2-0.069B^3)(X_t+97.69)=(1-0.919B)a_t\ \sigma^2_a=4418.18$$

#### Seasonal

First we read in the data collected from tswge.
```{r read in nottem}
x = data("nottem")
```

Convert the data to time series
```{r convert nottem to TS}
x=ts(nottem)
```

Plot the data and see seasonality in the sample autocorrelations
```{r seas Plot}
plotts.sample.wge(x)
```

The behavior in the realization tends to repeat every 12th (s) time periods.

The sample autocorrelations large at s12.

A seasonal model with season s=12 make sense.

Estimate the parameters of the "stationarized data".
```{r B12 overfit}
# overfit factor table of 24
over.x=est.ar.wge(x,p=24, factor = TRUE,type = 'burg')
```
Compare factors in overfit factor table with those for $(1 - B^s)$.
```{r B12 Factor Table}
factor.wge(p=c(0,0,0,0,0,0,0,0,0,0,0,1))
```

There is a somewhat close match so we will transform the data.

**Transformed data by $(1-B^{12})$**

```{r Transform B12}
# Transform differenced data with "seasonal difference"
y=artrans.wge(x,phi.tr=c(0,0,0,0,0,0,0,0,0,0,0,1))
```

The data looks somewhat stationary.
The next step is to model the transformed data with a $(1-B^{12})$
```{r B12 WNV}
# find the WNV
wnv=sd(y)^2
```
** Final Model**
$$(1-B^{12})X_t=a_t\ \sigma^2_a=11.82$$

#### Signal Plus Noise with Cochrane-Orcutt
First we need to install the correct libraries.
```{r install orcutt}
#install.packages("orcutt")
library(orcutt)
```
We will need to import the data and set up a data frame.
```{r read in orcutt}
# read in the data
x = read.csv("energydata_complete.csv", header = TRUE)
```

Convert the data to time series
```{r convert to orcutt TS}
#Just looking at the Appliances data
x=ts(x$Appliances)
```
Setup a data frame.
```{r data frame}
t = seq(1,19735,1)
df = data.frame(x = x, t= t)
```

Check the fit for OLS
```{r}
fit = lm(x~t, data = df)
summary(fit)
```
Check the fit for Cochrane Orcutt
```{r}
cfit = cochrane.orcutt(fit)
summary(cfit)
```
