---
title: "Unit 12 For Live Session"
author: "Chad Madding"
date: "March 22, 2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(GGally)
library(astsa)
library(tswge)
library(vars)
```

## Question Of Interest

The data set (LA_Cmort_Study.csv) is a portion of the data taken from a study (Shumway 1988) on the possible effects of pollution and temperature on weekly cardiac mortality (heart attacks) in Los Angeles County.

**Your goal is to utilize all given information to provide the most useful forecasts for the next 20 weeks of cardiac mortality.** 

You should include plots, tables, and charts to help make your analysis and inferences clear to your peers.

Start by reading in the data and see what it looks like.
```{r Read In LA Data}
CM = read.csv("la_cmort_study.csv", header = TRUE)
#Look at the first few rows
head(CM)
```

## Visualaze The Data

Make a matrix of plots with the LA_Cmort data set.

```{r pressure, echo=FALSE}
ggpairs(CM[2:4]) #matrix of scatter plots
```

#### ARIMA 1 MLR with Cor Errors (no lag, no seasonl categorical variable)
First we will forecast the Particles (part).
The $freq show a peek around 0.0192 showing some sign of a 52 week annual trend.
```{r forecast Particles, echo=TRUE, results='hide'}
#plot just the Particles (part)
plotts.sample.wge(CM$part) #peek in freq near .0192/1=52.08 (annual)
```

Using multivariate to forecast  
```{r}
CMnoX=CM[2:4]
#install.packages("vars")
# VAR and VARselect are from CRAN package vars
VARselect(CMnoX, lag.max = 15, type = "const",season = NULL, exogen = NULL)
#VARselect picks p=2 (using BIC)
lsfit=VAR(CMnoX,p=2,type="const")
preds=predict(lsfit,n.ahead=20)
# preds$fcst$ are the VAR forecasts for x1.  Similar for x2
#this will list out just the forcast
preds$fcst$temp
```


Transform the Particles (part) to remove the seasonality.
```{r ,echo=TRUE, results='hide'}
CM_52 = artrans.wge(CM$part, c(rep(0,51),1))
plotts.sample.wge(CM_52) #looks like some low freq?
aic5.wge(CM_52) #picks ARMA(2,1) assume stationary
aic5.wge(CM_52,type = "bic") #picks ARMA(2,1) 
ljung.wge(CM_52)$pval #FTR Ho
ljung.wge(CM_52, K = 48)$pval #FTR Ho
#Going with white noise despite peak at 0 in Spec D. 
#est = est.arma.wge(CM_52, p = 3, q = 2)
#CM_52_AR2_MA1 = artrans.wge(CM_52,est$phi)
predsPart = fore.aruma.wge(CM$part,s = 52, n.ahead = 20, limits = F)
```

Next we will forecast temperature (Temp).
```{r ,echo=TRUE, results='hide'}
#forecast Temp
plotts.sample.wge(CM$temp) #freq near .0192 (annual)
CM_52 = artrans.wge(CM$temp, c(rep(0,51),1))
plotts.sample.wge(CM_52) #looks like some low freq?
aic5.wge(CM_52) #picks ARMA(0,0)
aic5.wge(CM_52,type = "bic") #picks ARMA(0,0) 
ljung.wge(CM_52)$pval
ljung.wge(CM_52, K = 48)$pval #barely rejects
acf(CM_52,lag.max = 48) # acf looks consistent with white noise
predsTemp = fore.aruma.wge(CM$temp,s = 52, n.ahead = 20, limits = F)
```

Looking at cardiac mortality (heart attacks) in Los Angeles County based on temperature, particles and Week.
```{r}
# Model cmort based on predicted part and temp using MLR with Cor Erros
ksfit = lm(cmort~temp+part+Week, data = CM)
phi = aic.wge(ksfit$residuals)
```

Fit an ARIMA with the phi from above (2). Binding the temperature, particle and week data with cbind.
```{r}
fit = arima(CM$cmort,order = c(phi$p,0,0), seasonal = list(order = c(1,0,0), period = 52), xreg = cbind(CM$temp, CM$part, CM$Week))
```

Visually the acf is showing white noise. Now we need to check for whiteness of residuals with the Ljung-Box Test.
```{r}
#First visualize the residuals
acf(fit$residuals)
#Run the test with the default 24 and 48 maximum lag for sample autocorrelations to be used.
ljung.wge(fit$residuals) # pval = .048
ljung.wge(fit$residuals, K = 48) # pval = .002
```

Load the forecast Part and Temp data in a data frame. This will extend the week from 509 to 528 with **seq**.
```{r}
next20 = data.frame(temp = predsTemp$f, part = predsPart$f, Week = seq(509,528,1))
```

This will get the predictions
```{r}
#get predictions
predsCMort = predict(fit,newxreg = next20)
```

plot next 20 cmort wrt time
```{r}
plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,508), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(509,528,1), predsCMort$pred, type = "l", col = "red")
```

```{r ,echo=TRUE, results='hide'}
#Find ASE  Need to forecast last 30 of known series using dplyr lag.  
CMsmall = CM[1:478,]
CMsmall$temp_1 = dplyr::lag(CMsmall$temp,1)
CM$temp_1 = dplyr::lag(CM$temp,1)
ksfit = lm(cmort~temp_1+part+Week, data = CMsmall)
phi = aic.wge(ksfit$residuals)

fit = arima(CMsmall$cmort,order = c(phi$p,0,0), seasonal = list(order = c(1,0,0), period = 52), xreg = cbind(CMsmall$temp, CMsmall$part, CMsmall$Week))

last30 = data.frame(temp = CM$temp_1[479:508], part = CM$part[479:508], Week = seq(479,508,1))
#get predictions
predsCMort = predict(fit,newxreg = last30)

plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,528), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(479,508,1), predsCMort$pred, type = "l", col = "red")


ASE_ARMA1 = mean((CM$cmort[479:508] - predsCMort$pred)^2)
ASE_ARMA1
```



####ARIMA2: attempt at categorical variable for week but arima takes only continuous variables  

forecast Particles
```{r ,echo=TRUE, results='hide'}
CM = read.csv("la_cmort_study.csv", header = TRUE)
plotts.sample.wge(CM$part) #freq near .0192 (annual)
CM_52 = artrans.wge(CM$part, c(rep(0,51),1))
plotts.sample.wge(CM_52) #looks like some low freq?
aic5.wge(CM_52) #picks ARMA(2,1) assume stationary
aic5.wge(CM_52,type = "bic") #picks ARMA(0,0) 
ljung.wge(CM_52)$pval #FTR Ho
ljung.wge(CM_52, K = 48)$pval #FTR Ho
#Going with white noise despite peak at 0 in Spec D. 
#est = est.arma.wge(CM_52, p = 3, q = 2)
#CM_52_AR2_MA1 = artrans.wge(CM_52,est$phi)
predsPart = fore.aruma.wge(CM$part,s = 52, n.ahead = 20)
```

forecast Temp
```{r ,echo=TRUE, results='hide'}
plotts.sample.wge(CM$temp) #freq near .0192 (annual)
CM_52 = artrans.wge(CM$temp, c(rep(0,51),1))
plotts.sample.wge(CM_52) #looks like some low freq?
aic5.wge(CM_52) #picks ARMA(0,0)
aic5.wge(CM_52,type = "bic") #picks ARMA(0,0) 
ljung.wge(CM_52)$pval
ljung.wge(CM_52, K = 48)$pval #barely rejects
acf(CM_52,lag.max = 48) # acf looks consistent with white noise
predsTemp = fore.aruma.wge(CM$temp,s = 52, n.ahead = 20, limits = F)
```

Model **cmort** based on predicted **part** and **temp** using Multiple Linear Regression (MLR) with correlation errors.
```{r}
#This will add FWeek (52 week count) to the CM dataset
CM$FWeek = as.factor(CM$Week%%52)
ksfit = lm(cmort~temp+part+Week+FWeek, data = CM)
phi = aic.wge(ksfit$residuals)
fit = arima(CM$cmort,order = c(phi$p,0,0), xreg = cbind(CM$temp, CM$part, CM$Week, CM$FWeek))
```

Check for whiteness of residuals with a visual test and the Ljung-Box Test
```{r}
acf(fit$residuals)
ljung.wge(fit$residuals) # pval = .066
ljung.wge(fit$residuals, K = 48) # pval = .0058
```

load the forecasted Part and Temp in a data frame
```{r}
next20 = data.frame(temp = predsTemp$f, part = predsPart$f, Week = seq(509,528,1), FWeek = as.factor(seq(509,528,1)%%52))
```

get predictions - predict residuals   
```{r ,echo=TRUE, results='hide'}
plotts.sample.wge(ksfit$residuals)
phi = aic.wge(ksfit$residuals)
resids = fore.arma.wge(ksfit$residuals,phi = phi$phi,n.ahead = 20, limits = F)
#predict trend manually
preds = predict(ksfit, newdata = next20)

predsFinal = preds + resids$f
```


Plot the next 20 cmort wrt time
```{r}
plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,528), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(509,528,1), predsFinal, type = "l", col = "red")
length(predsFinal)
```

To Find the ASE we need to forecast last 30 of the known series. 
```{r}
CMsmall = CM[2:478,]
ksfit = lm(cmort~temp+part+Week+FWeek, data = CMsmall)
phi = aic.wge(ksfit$residuals)
fit = arima(CMsmall$cmort,order = c(phi$p,0,0), seasonal = list(order = c(1,0,0), period = 52), xreg = cbind(CMsmall$temp1, CMsmall$part, CMsmall$Week, CMsmall$FWeek))

last30 = data.frame(temp = CM$temp[479:508], part = CM$part[479:508], Week = seq(479,508,1), FWeek = as.factor(seq(479,508,1)%%52))
```

Predict residuals manually  
```{r}
plotts.sample.wge(ksfit$residuals)
phi = aic.wge(ksfit$residuals)
resids = fore.arma.wge(ksfit$residuals,phi = phi$phi,n.ahead = 30, limits = F)
```

Predict trend manually  
```{r}
preds = predict(ksfit, newdata = last30)
#Final using the trend and residuals
predsFinal = preds + resids$f
```

Plot and print out the ASE
```{r}
plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,528), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(479,508,1), predsFinal, type = "l", col = "red")

ASE_ARMA2 = mean((CM$cmort[479:508] - predsFinal)^2,na.rm = TRUE)
ASE_ARMA2
```

#### VAR Model

**VAR Model Forecasts Seasonally Differenced Data**  

Difference all series to make them stationary which is an assumption of VAR.  
They do not have to be white, just stationary.  

```{r}
CM = read.csv("la_cmort_study.csv", header = TRUE)
CM_52 = artrans.wge(CM$cmort,c(rep(0,51),1))
Part_52 = artrans.wge(CM$part,c(rep(0,51),1))
Temp_52 = artrans.wge(CM$temp,c(rep(0,51),1))

#VARSelect on Differenced Data chooses 2
VARselect(cbind(CM_52, Part_52, Temp_52),lag.max = 10, type = "both")

#VAR with p = 2
CMortDiffVAR = VAR(cbind(CM_52, Part_52, Temp_52),type = "both",p = 2)
preds=predict(CMortDiffVAR,n.ahead=20)

#We have predicted differences .... calculate actual cardiac mortalities 
startingPoints = CM$cmort[428:457]
CMortForcasts = preds$fcst$CM_52[,1:3] + startingPoints

#Plot
while (!is.null(dev.list()))  dev.off()
plot(seq(1,508,1), CM$cmort, type = "l",xlim = c(0,528), ylab = "Cardiac Mortality", main = "20 Week Cardiac Mortality Forecast")
lines(seq(509,528,1), as.data.frame(CMortForcasts)$fcst, type = "l", col = "red")

ASE_VAR1 = mean((CM$cmort[489:508] - CMortForcasts[,1])^2)
ASE_VAR1
```

### Modeling with the sales data set  

find the ASE for forecasts of the last five observations of the sales data set example for both the MLE model (using ARIMA) and the VAR model we fit.  

We fit several models using the arima() function, you should pick the one you feel is the best model (maybe with the lowest AIC?).

Your slides should include each of the following:
  
  * Your code  
  * Enough visual aids to describe how you calculated the ASE for each model  
  * The ASE for each model  
  * Which model you feel is better in that respect, and why.  

Read in and visualize the sales data  
```{r ,echo=TRUE, results='hide'}
BSales = read.csv("businesssales.csv", header = TRUE)

## Visualaze The Data
head(BSales)
#Make a matrix of plots with the BSales data set.
ggpairs(BSales[2:5]) #matrix of scatter plots
```

#### MLR Modeling

**Model 1**

All data with no lag and no trend
```{r}
ksfit=lm(sales~ad_tv+ad_online+discount, data = BSales)
aic.wge(ksfit$residuals,p=0:8, q=0)  # AIC picks p=7
fit=arima(BSales$sales,order=c(7,0,0),xreg=BSales[,3:5])
fit

acf(fit$residuals)
ltest = ljung.wge(fit$resid)
ltest$pval
```

ASE for model with no lag and no trend (last 5)
```{r}
#Cut off the last 5
BSales2 = BSales[1:95,]
ksfit=lm(sales~ad_tv+ad_online+discount, data = BSales2)
aic.wge(ksfit$residuals,p=0:8, q=0)  # AIC picks p=7
fit=arima(BSales2$sales,order=c(7,0,0),xreg=cbind(BSales2$ad_tv,BSales2$ad_online,BSales2$discount))
fit

preds = predict(fit, newxreg = cbind(BSales$ad_tv[96:100],BSales$ad_online[96:100],BSales$discount[96:100]))
ASE1 = mean((BSales$sales[96:100] - preds$pred)^2)
ASE1

while (!is.null(dev.list()))  dev.off()

plot(seq(1,100,1), BSales$sales[1:100], type = "l",xlim = c(0,100), ylab = "Business Sales", main = "5 Week Sales Forecast")
lines(seq(96,100,1), preds$pred, type = "l", col = "red")
```

**Model 2**

ASE for model with no lag and trend (last 5)  
```{r}
BSales = read.csv("businesssales.csv", header = TRUE)
#Create trend (length of the dataset)
t=1:100
#add trend to the BSales dataset
BSales$t = t
#Take off the last 5 for testing the forcast
BSales2 = BSales[1:95,]
#Trend (t) has been added to the model
ksfit=lm(sales~t+ad_tv+ad_online+discount, data = BSales2)
aic.wge(ksfit$residuals,p=0:8, q=0)  # AIC picks p=6
fit=arima(BSales2$sales,order=c(6,0,0),xreg=cbind(BSales2$ad_tv,BSales2$ad_online,BSales2$t,BSales2$discount))
fit

preds = predict(fit, newxreg = cbind(BSales$ad_tv[96:100],BSales$ad_online[96:100],BSales$t[96:100],BSales$discount[96:100]))
ASE2 = mean((BSales$sales[96:100] - preds$pred)^2)
ASE2

while (!is.null(dev.list()))  dev.off()
plot(seq(1,100,1), BSales$sales[1:100], type = "l",xlim = c(0,100), ylab = "Business Sales", main = "5 Week Sales Forecast")
lines(seq(96,100,1), preds$pred, type = "l", col = "red")
```

```{r}
#Lagging Variables
#Example:
#With dplyr lag function
library(dplyr)
df = data.frame(Y = c(1,1,2,3,4,4,5,8),X1 = c(5,6,6,7,7,8,8,9))
df$X1_L1 = dplyr::lag(df$X1,1)
df$X1_L2 = dplyr::lag(df$X1,2)
df

# Model 3

#Lagging BSales Ad Variables
ad_tv1 = dplyr::lag(BSales$ad_tv,1)
ad_online1 = dplyr::lag(BSales$ad_online,1)
BSales$ad_tv1= ad_tv1
BSales$ad_online1 = ad_online1


#with trend and lagging

# ASE for model with no lag and trend (last 5)
t=1:100
BSales$t = t
BSales2 = BSales[2:95,]
ksfit=lm(sales~t+ad_tv1+ad_online1+discount, data = BSales2)
aic.wge(ksfit$residuals,p=0:8,q=0:0)  # AIC picks p=7
fit = arima(BSales2$sales,order = c(7,0,0), xreg = cbind(BSales2$ad_tv1,BSales2$ad_online1,BSales2$t,BSales2$discount))
fit


preds = predict(fit, newxreg = cbind(BSales$ad_tv1[96:100],BSales$ad_online1[96:100],BSales$t[96:100],BSales$discount[96:100]))
ASE3 = mean((BSales$sales[96:100] - preds$pred)^2)
ASE3


while (!is.null(dev.list()))  dev.off()
plot(seq(1,100,1), BSales$sales[1:100], type = "l",xlim = c(0,100), ylab = "Business Sales", main = "5 Week Sales Forecast")
lines(seq(96,100,1), preds$pred, type = "l", col = "red")





####### Forecast Features 

plotts.sample.wge(BSales$ad_tv)
aic5.wge(BSales$ad_tv)
est_ad_tv = est.arma.wge(BSales$ad_tv,p = 2, q = 2)
ad_tvFORECAST = fore.arma.wge(BSales$ad_tv,phi = est_ad_tv$phi, theta = est_ad_tv$theta, n.ahead = 6)

plotts.sample.wge(BSales$ad_online)
aic5.wge(BSales$ad_online, p = 0:10)
est_online = est.arma.wge(BSales$ad_online,p = 6)
while (!is.null(dev.list()))  dev.off()
plot.ts(BSales$ad_online[1:100])
ad_onlineFORECAST = fore.arma.wge(BSales$ad_online,phi = est_online$phi, n.ahead = 6)



#with trend and lagging

ad_tvFORECAST1 = lag(ad_tvFORECAST,1)

ad_onlineFORECAST1 = lag(ad_onlineFORECAST,1)

# ASE for model with no lag and trend (last 5)
t=1:100
BSales$t = t
BSales2 = BSales[2:95,]
ksfit=lm(sales~t+ad_tv1+ad_online1, data = BSales2)
aic.wge(ksfit$residuals,p=0:8,q=0:0)  # AIC picks p=7
fit = arima(BSales2$sales,order = c(7,0,0), xreg = cbind(BSales2$ad_tv1,BSales2$ad_online1,BSales2$t))
fit


preds = predict(fit, newxreg = cbind(ad_tvFORECAST$f[2:6],ad_onlineFORECAST$f[2:6],BSales$t[96:100]))
ASE3.5 = mean((BSales$sales[96:100] - preds$pred[1:5])^2)
ASE3.5


plot(seq(1,100,1), BSales$sales[1:100], type = "l",xlim = c(0,100), ylab = "Business Sales", main = "5 Week Sales Forecast")
lines(seq(96,100,1), preds$pred[2:6], type = "l", col = "red")
```



####VAR MODELS

**Model 4**

```{r}

BSVar = VAR(cbind(BSales2$sales,BSales2$ad_tv1,BSales2$ad_online1), type = "both", lag.max = 10)

preds = predict(BSVar,n.ahead = 5)
                  
ASE4 = mean((BSales$sales[96:100] - preds$fcst$y1[,1])^2)
ASE4

while (!is.null(dev.list()))  dev.off()
plot(seq(1,100,1), BSales$sales[1:100], type = "l",xlim = c(0,100), ylab = "Business Sales", main = "5 Week Sales Forecast")
lines(seq(96,100,1), preds$fcst$y1[,1], type = "l", col = "red")
```

**Model 5**

```{r}
BSVar = VAR(cbind(BSales2$sales,BSales2$ad_tv,BSales2$ad_online), type = "both", lag.max = 10)

preds = predict(BSVar,n.ahead = 5)

ASE5 = mean((BSales$sales[96:100] - preds$fcst$y1[,1])^2)
ASE5

while (!is.null(dev.list()))  dev.off()
plot(seq(1,100,1), BSales$sales[1:100], type = "l",xlim = c(0,100), ylab = "Business Sales", main = "5 Week Sales Forecast")
lines(seq(96,100,1), preds$fcst$y1[,1], type = "l", col = "red")
```




